---
title: "The attest contract: what a certificate guarantees, and what it does not"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The attest contract: what a certificate guarantees, and what it does not}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(42)
```

## The gap

A fitted model in R is an object that will answer any question you put to it.
Hand `predict()` a row from a population the model has never seen, a row with a
feature ten standard deviations outside the training range, or a row that has
drifted for six months since the model was fitted, and you get a number back.
The number carries no indication that it should not be trusted.

The validation you did lives somewhere else: in a script, a notebook, a slide,
a memory. Nothing binds it to the object. The object cannot tell you whether it
was ever checked for leakage, whether its probabilities were calibrated, or on
what data the claim was made. By the time the model is in production, the
evidence and the artefact have parted company.

**attest** closes that gap by making the evidence part of the object and the
refusal part of the prediction path.

```{r}
library(attest)
```

## The contract

An `attested_model` upholds four clauses.

1. **Checks run at fit time, not on request.** Leakage, calibration, coverage
   and shift checks are executed by `attest_fit()` itself. There is no code
   path that fits a model without running them.

2. **Results are stored inside the object and bound by hash.** The certificate
   records every check, its statistic, its threshold, and SHA-256 hashes of the
   training features, the specification and the fitted model.

3. **Predictions carry a validity status.** Every row comes back with an
   interval or label set, a `.status`, and a `.reason` when that status is not
   `valid`.

4. **Refusal is the default.** Rows outside the certified support get `NA`, not
   a number. Overriding that requires `enforce = FALSE`, and the override is
   recorded on the returned object.

The unit of configuration is the *specification*:

```{r}
spec <- attest_spec(split = split_random(prop = 0.2, calib = 0.2))
spec
```

`default_checks()` supplies the standard set. `on_fail` decides what a blocking
failure means: `"refuse"` (the default: no certificate, `attest_fit()` errors),
`"flag"` or `"warn"`.

## A certificate that issues

```{r}
n <- 4000
d <- data.frame(x1 = rnorm(n), x2 = rnorm(n), x3 = runif(n, 0, 10))
d$y <- factor(rbinom(n, 1, plogis(0.8 * d$x1 - 0.5 * d$x2)))

m <- attest_fit(spec, y ~ x1 + x2 + x3, d, engine_glm())
```

The checks report as they run, and a certificate is issued. It is worth
noticing that `x3` is pure noise: attest makes no claim that the model is
*good*, only that the stated checks were run and passed.

```{r}
certificate(m)
```

## A certificate that is refused

Now the case the package exists for. We add `score`, a feature that is almost a
restatement of the outcome -- the classic shape of a leak, where a field
recorded downstream of the label finds its way into the training frame.

```{r}
leaky <- d
leaky$score <- ifelse(leaky$y == "1", rnorm(n, 3), rnorm(n, -3))
```

```{r error = TRUE}
attest_fit(spec, y ~ x1 + x2 + x3 + score, leaky, engine_glm(), quiet = TRUE)
```

No object is returned. The fit does not complete.

This is the point of the package, so it is worth being precise about what just
happened. Fit that same leaky model without the blocking check and inspect it
the way a careful analyst would:

```{r warning = FALSE}
peek <- attest_fit(
  attest_spec(on_fail = "flag"), y ~ x1 + x2 + x3 + score, leaky,
  engine_glm(), quiet = TRUE
)
compare <- function(x) {
  r <- certificate(x)$results
  c(ECE = r$calib_ece$statistic,
    coverage = r$conformal_split$statistic,
    q = r$conformal_split$evidence$q)
}
round(rbind(honest = compare(m), leaky = compare(peek)), 4)
```

Read that table carefully, because it is the argument for the whole package.

The leaky model is **better calibrated** by a factor of about thirty, and its
conformal quantile `q` collapses to zero -- meaning its label sets are
singletons where the honest model mostly returns the uninformative `{0,1}`. On
the two properties a reviewer actually looks at, calibration and sharpness, the
leaking model is strictly better. Leakage does not look like a problem in the
metrics; it looks like success.

Coverage, meanwhile, is near 0.89 for both, and that is neither a coincidence
nor a defect. Split conformal attains its target coverage *by construction*,
for any underlying model, however good or bad it is. Coverage is a validity
property of the procedure, not a quality signal about the model. It cannot tell
a leaking model from an honest one, and it was never able to.

So neither calibration nor coverage will save you here: one is fooled, the
other is blind. That is why the leakage check has to run automatically, at fit
time, and block -- rather than being one more diagnostic a busy analyst is
invited to consult.

Note also that `on_fail = "flag"` issues the certificate but marks it `failed`,
and a model whose certificate is not `valid` is not sealed, so `predict()` will
refuse it. Flagging lets you *inspect* a failing model; it does not let you use
one.

### Waivers are loud

Sometimes a check is wrong about your data. You can waive it, but not silently:
a waiver requires a reason, and that reason is written into the certificate and
reprinted in every report thereafter.

```{r}
waived <- attest_fit(
  spec, y ~ x1 + x2 + x3 + score, leaky, engine_glm(),
  waive = "leak_target_proxy",
  reason = "score is a validated underwriting rating available before the claim",
  quiet = TRUE
)
certificate(waived)$waivers
```

The design intent is that a waiver is an on-the-record engineering decision
rather than a way to make a warning disappear.

## Predictions carry status

```{r}
predict(m, d[1:5, ])
```

`.set` is the conformal label set. A `{0,1}` row is not a failure -- it is the
model declining to distinguish the classes at the requested confidence level.
An empty set `{}` is also possible, and means no label is plausible at that
level: a strong signal that the row does not belong.

Now feed it rows outside the training support:

```{r}
nd <- d[1:5, ]
nd$x1 <- c(0.2, 12, -11, 0.5, 0)
predict(m, nd)
```

Rows 2 and 3 are refused. To see why that matters, override the contract:

```{r}
predict(m, nd, enforce = FALSE)
```

The refused rows come back maximally confident -- probabilities pinned near 1
and near 0 -- on inputs the model has no evidence about. Extrapolation does not
announce itself as uncertainty; it announces itself as certainty. Note that the
object records `(NOT enforced)`, so the override travels with the result.

### Shift is a batch property

Support is checked per row. Distribution shift is a property of a batch, so it
needs enough rows to estimate (`min_batch`, default 50):

```{r}
big <- d[1:200, ]
big$x1 <- big$x1 + 2
pb <- predict(m, big)
table(pb$.status)
unique(pb$.reason[pb$.status == "flagged"])
```

Most rows are individually inside the training range, so they are not refused --
but the *batch* has moved, and PSI catches it. Flagged means the prediction is
returned and the caller is told the population has changed.

## Checks know how uncertain they are

Every check statistic above is estimated from a finite sample, so comparing a
point estimate to a threshold decides borderline cases by luck. Checks
therefore report a bootstrap confidence interval, and the verdict follows a
three-way rule:

- **fail** -- the whole interval clears the threshold. The violation is
  confident.
- **pass** -- the whole interval is on the acceptable side.
- **weak** -- the interval straddles the threshold. The evidence does not
  settle it.

```{r}
attest_verdict(0.20, c(0.15, 0.25), threshold = 0.10, direction = "at_most")
attest_verdict(0.04, c(0.02, 0.06), threshold = 0.10, direction = "at_most")
attest_verdict(0.09, c(0.04, 0.15), threshold = 0.10, direction = "at_most")
```

A `weak` check passes by default, on the principle that noise alone should not
fail a model. But it is reported, stored in the certificate and printed in the
model card, so "we could not tell" is never silently recorded as "fine":

```{r}
mid <- mean(certificate(m)$results$calib_ece$ci)
borderline <- attest_spec(checks = list(calib_ece(max = mid), shift_monitor()))
mw <- attest_fit(borderline, y ~ x1 + x2 + x3, d, engine_glm())
```

This cuts against fail-closed defaults in one specific way, worth stating
plainly: intervals widen as data gets scarcer, so the smaller your dataset, the
harder it is for any check to fail. A package that refuses by default would
become most permissive exactly when evidence is thinnest. `strict = TRUE`
inverts that, treating an inconclusive check as a failure:

```{r error = TRUE}
attest_fit(
  attest_spec(checks = list(calib_ece(max = mid), shift_monitor()), strict = TRUE),
  y ~ x1 + x2 + x3, d, engine_glm(), quiet = TRUE
)
```

Not every check gets an interval. `leak_duplicates()` reports none, because
duplication is a census of the rows you hold rather than an estimate of a
population quantity: if a training row is in your test set, it is there, and
resampling would only describe some other dataset.

PSI is a third case again. Its noise floor depends on the batch size, so
resampling the batch would not help -- the bootstrap would inherit the same
small-sample bias. With ten bins, a batch of 50 rows drawn from the *training
distribution itself* has a median PSI of about 0.2, which is the conventional
flag threshold: the default settings would flag roughly half of all healthy
small batches. So the null is simulated instead, drawing batches of the
observed size from the stored training frequencies, and a feature is flagged
only when it beats both that null and the effect-size threshold.

## The model card

`report()` renders Markdown entirely from the certificate, so it cannot drift
from the model it describes:

```{r}
cat(report(m)[1:12], sep = "\n")
```

## What a certificate does not guarantee

This section matters more than the rest of the vignette. A certificate is
evidence that stated checks were run and passed on stated data. It is not a
warranty, and reading it as one would be worse than having no certificate at
all.

**Conformal coverage is marginal, not conditional.** `alpha = 0.1` gives
intervals covering the truth about 90% of the time *averaged over the
population*. It says nothing about any particular row or subgroup. Coverage can
be 90% overall while being 99% for the bulk of your data and 40% for a minority
segment. If you need per-group validity, check it per group; attest does not do
this for you.

**Coverage is finite-sample and random.** Split conformal guarantees
`1 - alpha` coverage in expectation over draws of the calibration set, within
about `1 / (n_calib + 1)`. Empirical coverage on any one test set scatters
around the target, and a reading slightly below `1 - alpha` is sampling noise
rather than a broken guarantee. The `tolerance` argument to `conformal_split()`
exists for exactly this reason.

**Coverage is not a measure of model quality.** As the leakage example above
shows, a conformal procedure hits its target coverage whatever model it wraps.
A useless model achieves 90% coverage by returning enormous intervals or label
sets containing every class. Coverage tells you the procedure is valid; only
sharpness -- interval width, or the quantile `q` -- tells you it is useful. Read
the two together or neither means anything.

**Exchangeability is the whole assumption, and shift breaks it.** The conformal
guarantee holds when calibration and future data are exchangeable. Under
distribution shift it is simply void -- not degraded in a quantified way, void.
This is the honest reason attest refuses out-of-support rows: refusal is not an
extra safety feature layered on top of the guarantee, it is an admission that
outside the training support **there is no guarantee left to offer**. The
refusal and the interval are two halves of one claim.

**Support is checked marginally, not jointly.** A row is refused when a feature
falls outside its own training range. A row can sit comfortably inside every
feature's range and still be far from the joint distribution -- a 40-year-old
with a 45-year career. Marginal support checks will not catch that, and the
`.shift` score will not either.

**PSI is a heuristic, and its threshold is an effect size.** The 0.2 line is
convention, not theory. The batch size problem is handled -- see *Checks know
how uncertain they are* above -- but the rest is not: PSI is computed per
feature on binned marginals, so it will miss a shift in the dependence
structure between features that leaves every marginal intact, and it will fire
on harmless seasonal movement in a feature the model barely uses. Because
`threshold` acts as an effect-size floor, a real but subtle shift is never flagged
however many rows you have. Treat a flag as a prompt to investigate, not a
verdict.

**The leakage checks are shallow by construction.** `leak_target_proxy()` fits
one-variable models, so it finds single features that restate the target. It
will not find a leak spread across two features, nor one mediated by a grouping
structure. `leak_duplicates()` finds exact duplicate rows, not near-duplicates.
Passing means the obvious leaks are absent, not that the data is clean.

**Calibration is measured where it was measured.** ECE is computed on the test
split at fit time, with equal-width bins. It is a statement about that data on
that day. Nothing recomputes it later.

**Hashes prove integrity, not quality.** `verify()` confirms the model object
has not been altered since certification. It does not, and cannot, confirm the
training data was appropriate, correctly labelled, or ethically obtained.
`verify(x, data)` currently checks only that the expected feature columns are
present, because the split cannot be reconstructed without the original seed.

**A certificate has a shelf life.** It is a claim about a moment. Nothing in
the object expires it, and a nine-month-old certificate on a drifted population
is a statement about a world that no longer exists.

The scope of the claim is deliberately narrow: *these checks, this data, this
moment*. Its value comes from being immutable, machine-readable and attached to
the artefact -- not from being broad.

## Extending

Checks are the extension point. A check is a function of the fitting context
returning an `attest_result()`:

```{r}
max_missing <- function(max_prop = 0.05) {
  new_check("max_missing", stage = "pre", blocking = TRUE, run = function(ctx) {
    prop <- mean(is.na(ctx$train[ctx$features]))
    attest_result(
      "max_missing",
      if (prop <= max_prop) "pass" else "fail",
      statistic = prop, threshold = max_prop,
      message = sprintf("%.1f%% of training feature cells missing", 100 * prop)
    )
  })
}

custom <- attest_spec(checks = c(default_checks(), list(max_missing())))
m2 <- attest_fit(custom, y ~ x1 + x2 + x3, d, engine_glm(), quiet = TRUE)
certificate(m2)$results$max_missing
```

`stage = "pre"` runs before fitting and sees only data; `stage = "post"` runs
after and sees `ctx$model`. Setting `blocking = FALSE` makes a check
informational -- it is recorded in the certificate and appears in the report,
but never prevents issuance.
